---
title: "Assignment 02 Submission"
author: "Sean Kim"
format:
  html:
    embed-resources: true
---

```{r}
library(data.table)
library(dplyr)
library(tidyverse)
library(leaflet)

if (!file.exists("01_chs_individual")) 
  download.file( 
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/01_chs/chs_individual.csv", 
    destfile = "01_chs_individual", 
    method = "libcurl", 
    timeout = 1000 )


ind <- fread("01_chs_individual") 



if (!file.exists("01_chs_regional")) 
  download.file( 
    url = "https://raw.githubusercontent.com/USCbiostats/data-science-data/master/01_chs/chs_regional.csv", 
    destfile = "01_chs_regional", 
    method = "libcurl", 
    timeout = 1000 )
reg <- fread("01_chs_regional")
```

```{r}
reg[, n:= 1:.N, by = .(townname)]
reg <- reg[n == 1,][, n := NULL]

merged_data <- inner_join(ind, reg, by = "townname")
duplicate_count <- sum(duplicated(merged_data))
cat("Number of duplicate rows:", duplicate_count, "\n")
merged_data <- merged_data[!duplicated(merged_data), ]

```

# Data Wrangling

## Question 1 - Merge and Impute missing values

```{r}
#numeric missing values - replace with mean
merged_data <- merged_data %>%
  group_by(male, hispanic) %>%
  mutate(across(where(is.numeric), ~ifelse(is.na(.), mean(., na.rm = TRUE), .)))

#categorical missing values - replace with mode
merged_data <- merged_data %>%
  group_by(male, hispanic) %>%
  mutate(across(where(is.character), ~ifelse(is.na(.), Mode(.), .)))

```

## Question 2 - New variables

```{r}
merged_data <- merged_data %>%
  mutate(obesity_level = cut(bmi, breaks = c(-Inf, 14, 22, 24, Inf), labels = c("Underweight", "Normal", "Overweight", "Obese")))


summary_byObLvl <- merged_data %>%
  group_by(obesity_level) %>%
  summarise(
    min_BMI = min(bmi, na.rm = TRUE),
    max_BMI = max(bmi, na.rm = TRUE),
    total_observations = n()
  )
summary_byObLvl
```

## Question 3 - smoke gas exposure

```{r}
merged_data <- merged_data %>%
  mutate(smoke_gas_exposure = case_when(
    smoke == 0 & gasstove == 0 ~ "No Exposure",
    smoke > 0 & gasstove == 0 ~ "Smoke Exposure",
    smoke == 0 & gasstove > 0 ~ "Gas Exposure",
    smoke > 0 & gasstove > 0 ~ "Both"
  ))

head(merged_data)

```

## Question 4 - Summary tables of FEV1

```{r}
summary_table_town <- merged_data %>%
  group_by(townname) %>%
  summarise(
    avg_FEV = mean(fev, na.rm = TRUE),
    sd_FEV = sd(fev, na.rm = TRUE)
  )

summary_table_sex <- merged_data %>%
  group_by(male) %>%
  summarise(
    avg_FEV = mean(fev, na.rm = TRUE),
    sd_FEV = sd(fev, na.rm = TRUE)
  )

summary_table_obesity <- merged_data %>%
  group_by(obesity_level) %>%
  summarise(
    avg_FEV = mean(fev, na.rm = TRUE),
    sd_FEV = sd(fev, na.rm = TRUE)
  )

summary_table_exposure <- merged_data %>%
  group_by(smoke_gas_exposure) %>%
  summarise(
    avg_FEV = mean(fev, na.rm = TRUE),
    sd_FEV = sd(fev, na.rm = TRUE)
  )

print("Summary Table by Town:")
print(summary_table_town)

print("\nSummary Table by Sex:")
print(summary_table_sex)

print("\nSummary Table by Obesity Level:")
print(summary_table_obesity)

print("\nSummary Table by Smoke Gas Exposure:")
print(summary_table_exposure)
```

# EDA

## Question 1: BMI vs FEV?

```{r}
summary(merged_data$bmi)
summary(merged_data$fev)

cor(merged_data$bmi, merged_data$fev, use = "complete.obs")

ggplot(merged_data, aes(x = bmi, y = fev)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "blue") +
  labs(title = "Association between BMI and FEV",
       x = "BMI",
       y = "FEV")
```

Correlation between BMI and FEV = 0.357, indicating every one point increase in BMI is associated with a 0.357 unit increase in FEV.

## Question 2: smoke/gas exposure vs FEV?

```{r}
summary(merged_data$smoke_gas_exposure)
summary(merged_data$fev)

ggplot(merged_data, aes(x = smoke_gas_exposure, y = fev)) +
  geom_boxplot() +
  labs(title = "Association between Smoke/Gas Exposure and FEV",
       x = "Smoke/Gas Exposure",
       y = "FEV")
```

Exposure to both smoke and gas seems to be associated with a lower FEV compared to just one or the other. There is significant variation between groups, evidenced by overlap in the box plots.

## Question 3: PM2.5 vs FEV?

```{r}
summary(merged_data$pm2_5_fr)
summary(merged_data$fev)

cor(merged_data$pm2_5_fr, merged_data$fev, use = "complete.obs")

ggplot(merged_data, aes(x = pm2_5_fr, y = fev)) +
  geom_point() +
  geom_smooth(method = "lm", se = FALSE, color = "green") +
  labs(title = "Association between PM2.5 Exposure and FEV",
       x = "PM2.5",
       y = "FEV")
```

Slight negative association between PM2.5 concentration exposure and FEV (cor = -0.08). This is reasonable considering PM2.5 is a fine particle that would likely damage the lungs. Increased exposure to PM2.5 should lead to a lower FEV.

# Data Visualization

1.  Facet plot showing scatterplots with regression lines of BMI vs FEV by "townname".

    ```{r}
    ggplot(merged_data, aes(x = bmi, y = fev))+
             geom_point()+
             geom_smooth(method = "lm", se = FALSE, color = "blue") + 
             facet_wrap(~townname, scales = "free")+ 
             labs(title = "Scatterplots of BMI vs FEV by Town",
           x = "BMI",
           y = "FEV")
    ```

2.  Stacked histograms of FEV by BMI category and FEV by smoke/gas exposure. Use different color schemes than the ggplot default.

    ```{r}
    custom_colors <- c("Underweight" = "lightblue", "Normal" = "forestgreen", "Overweight" = "orange", "Obese" = "red")

    ggplot(merged_data, aes(x = fev, fill = obesity_level)) +
      geom_histogram(position = "stack", bins = 30, color = "white") +
      scale_fill_manual(values = custom_colors) +
      labs(title = "Stacked Histograms of FEV by BMI Category",
           x = "FEV",
           fill = "BMI cat") +
      facet_wrap(~smoke_gas_exposure, scales = "free")
    ```

3.  Barchart of BMI by smoke/gas exposure.

    ```{r}
    bmi_smokegas_data <- merged_data %>%
      group_by(smoke_gas_exposure) %>%
      summarize(mean_bmi = mean(bmi))
    bmi_smokegas_data

    ggplot(bmi_smokegas_data, aes(x = smoke_gas_exposure, y = mean_bmi, fill = smoke_gas_exposure)) +
      geom_bar(stat = "identity") +
      labs(title = "Bar Chart of BMI by Smoke/Gas Exposure",
           x = "Smoke/Gas Exposure",
           y = "BMI",
           fill = "Smoke/Gas Exposure")
    ```

4.  Statistical summary graphs of FEV by BMI and FEV by smoke/gas exposure category.

    ```{r}
    ggplot(merged_data, aes(x = bmi, y = fev)) + 
      geom_point()+
      geom_smooth(method = "lm", se = FALSE)+ 
      labs(title = "Scatterplot of FEV by BMI with Regression Lines by Smoke/Gas Exposure",
           x = "BMI",
           y = "FEV",
           color = "Smoke/Gas Exposure") +
      facet_wrap(~smoke_gas_exposure, scales = "free")
    ```

    Exposure to both smoking and gas stove seems to increase the effect of increasing BMI on increasing FEV.

5.  A leaflet map showing the concentrations of PM2.5 mass in each of the CHS communities.

    ```{r}
    leaflet(merged_data) %>%
      addTiles() %>%
      addCircles(lng = ~lon, lat = ~lat, weight = 5, radius = ~pm2_5_fr * 10, popup = ~paste(townname, "PM2.5: ", pm2_5_fr))
    ```

6.  Choose a visualization to examine whether PM2.5 mass is associated with FEV.

    ```{r}
    ggplot(merged_data, aes(x = pm2_5_fr, y = fev))+ 
      geom_point()+
      geom_smooth(method = "lm", se = FALSE)+ 
      labs(title = "Scatterplot of FEV by PM2.5 with Regression Line",
           x = "PM2.5",
           y = "FEV")


    ```

```{r}
merged_data %>% 
  mutate(pm2_5_fr = round(pm2_5_fr, 4))

merged_data$pm2_5_fr <- as.factor(merged_data$pm2_5_fr)
anova_result <- aov(fev ~ pm2_5_fr, data = merged_data)
summary(anova_result)

tukey_results <- TukeyHSD(anova_result)
summary(tukey_results)

ggplot(merged_data, aes(x = pm2_5_fr, y = fev, color = pm2_5_fr)) +
  geom_boxplot() +
  labs(title = "Boxplot of FEV by PM2.5 Levels",
       x = "PM2.5 Levels",
       y = "Forced Expiratory Volume (mL)")+
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
```

```{r}

FEV_PM2_5_groups <- merged_data %>% 
  group_by(pm2_5_fr) %>% 
  summarize(mean_fev = mean(fev))
FEV_PM2_5_groups

ggplot(FEV_PM2_5_groups, aes(x = pm2_5_fr, y = mean_fev, fill = pm2_5_fr)) + 
  geom_bar(stat = "identity") +
  labs(title = "FEV by 2.5 Levels", 
       x = "PM2.5 Level", 
       y = "FEV")+
   theme(axis.text.x = element_text(angle = 45, hjust = 1))
  
         
```
